A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay

### 2021-01-19 Paper/NatureAstronomy/main.tex

\documentclass[12pt]{article}

@@

%\usepackage{amssymb}
\usepackage{xspace}
\usepackage{graphicx}

@@

\usepackage{scicite}

@@

\paragraph*{One Sentence Summary:} A distant supernova has appeared as three images magnified by the gravity of a galaxy cluster, and will reappear in 20 years.

@@

  We report the discovery of a strongly-lensed supernova (SN) explosion that will enable a time delay measurement with an uncertainty of $<1\%$. This unprecedented scenario will enable a rare and highly precise measurement of the Hubble constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACS J0138.0-2155.  This is the most distant multiply-imaged supernova yet discovered.  We classify the supernova as a Type Ia with {$>90\%$} probability based on archival imaging observations from the \textit{Hubble Space Telescope}.  Three lensed images of the SN are detected with relative time delays of $<$200 days, and we predict that a fourth image in the component closest to the cluster core will appear in the year 2037$\pm$2.  Observing the light curve of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary 20 year baseline.

@@

\section*{Main Text:}

Observations of Type Ia supernova (SN) explosions have played a key role in mapping the cosmic expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate\cite{riess_observational_1998,perlmutter_measurements_1999, Riess_large_2019}.  Determining the nature of dark energy and how it may evolve over time is a primary goal for the large-scale cosmology experiments of the 2020s \cite{amendola_cosmology_2013,spergel_wide_2015,Ivezic_lsst_2019}.
A series of recent investigations into the expansion rate of the universe (the Hubble-LeM\^aitre constant; $H_0$) have found that measurements from the local universe are significantly different from the value inferred from measurement of the cosmic microwave background (CMB) radiation \cite{Riess_large_2019,aghanim_planck_2018}.  Resolving this tension in $H_0$ measurements may reveal new physics from the early universe, and to do so requires multiple 
independent cosmological probes \cite{verde_tensions_2019}.

One promising cosmological tool uses %time delays of 
gravitational lens systems in which a background
source appears as multiple images that arrive to the observer with relative delays \cite{einstein_lens_1936}.
%, which %for a given lens potential 
When such a strongly-lensed source is variable, one can measure the time delay between any pair of lensed images and derive a ratio of angular diameter distances to the foreground lens and background source. 
This distance ratio is sensitive to cosmological parameters, such as 
%the expansion rate of the universe, 
$H_0$  and the dark energy equation-of-state, $w$ \cite{refsdal_possibility_1964,coe_cosmological_2009,linder_lensing_2011}.
In recent years this method has been successfully applied to lensed
quasars \cite{suyu_cosmology_2014,bonvin_cosmograil_2019}, with six high-precision measurements to date \cite{wong_h0licow_2019}. As this sample of time delay lenses grows to several dozen, it is expected to deliver a measurement of $H_0$ with 1\% precision \cite{suyu_cosmological_2018}.

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---in contrast to the stochastic variation of quasars.  However, despite the promise of lensed SN time delay cosmography as a cosmological distance probe, to date, there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_sn_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2019}. Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

@@

Future large-scale surveys such as the Vera C. Rubin Observatory 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their mission lifetimes \cite{oguri_gravitationally_2010,goldstein_rates_2019,wojtak_magnified_2019}. However, many of these lensed SN will require substantial follow-up resources to be made cosmologically useful. Furthermore, almost all will have significantly shorter 
delays (of order 10--100 days) and therefore higher relative uncertainties than \SNABC.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like SN Requiem can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by the time delay measurement errors.

In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN.  For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of \SNABC, the constraints 
from the \reqlike sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \reqlike sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \reqlike contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.

@@

\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout_sn4.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the \SNABC images  (SN1--3) and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  
    %Labels marked 3.1--3.4 indicate the locations of a separate  multiply-imaged [O\textsc{ii}]-emitter at $z=0.77$ used to help constrain the lensing potential. 
    Panels \emph{b}--\emph{i} show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels \emph{bcde} show the imaging from 2016 July where the SN was visible and \emph{fghi} show the later imaging from 2019 July where the the SN has faded away.  Panels \emph{ei} indicate the location of the fourth image predicted to appear in $\approx$2037. All panels use the late-epoch F125W imaging for the green channel; nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which helps to constrain the relative age of each SN image and the transient classification as a likely Type Ia supernova explosion.}

    \label{fig:layout}
\end{figure*}
\clearpage
\begin{figure}
  \centering
    \includegraphics[draft=False, width=\textwidth]{Paper/Figures/classification_contours_timeline.pdf}
    \caption{Classification information for SN Requiem based on its position in color-magnitude space. \emph{Left:} Contours show the population distributions for normal SNe of Type Ia (red), Type Ib/Ic (gold), and Type II (green), drawn to enclose 68\% (interior solid lines) and 95\% (exterior shaded regions) of each SN population.  Each SN sub-class was simulated at $z=1.95$, and samples from their expected light curves were drawn uniformly in time. Open markers show the observed photometry of the SN. Dotted vertical lines mark the lensing magnification correction, terminating in closed markers and solid bars that indicate the magnification in our preferred lens model and its systematic uncertainty, respectively.  \emph{Right:} The evolution of a typical Type Ia SN at $z=1.95$ in observed color-brightness space is shown by a colored line, with the line color indicating SN age in observer-frame days relative to peak brightness.  White diamonds correspond to the times labeled on the colorbar at right. Grey shading shows the typical range of luminosities observed for the Type Ia SN population in the nearby universe \cite{wang_determination_2006}. 
    %An interactive 3-D version of this figure is available online at \href{https://plot.ly/~jpierel/6}{https://plot.ly/~jpierel/6}
    \label{fig:class}
    }
\end{figure}

@@

\midrule
\textit{1}  & $99^{+27}_{-29}$ & -- & $3.9\pm0.5$ & --\\
\textit{2} & $-19^{+21}_{-13}$ & $114^{+39}_{-42}$ & $7\pm3$ & $101\pm62$ \\
\textit{3} & $87^{+15}_{-13}$ & $11\pm23$ & $5\pm1$ & $19\pm34$ \\
\textit{4} & -- & -- & $0.4\pm0.2$ & $7760\pm540$\\

@@

%BibTeX users: After compilation, comment out the following two lines and paste in
% the generated .bbl file. 
%%%%%%%%%%%% Bibliography
%\bibliography{references}
%\bibliographystyle{Science}
\begin{thebibliography}{10}

\bibitem{riess_observational_1998}
A.~G. Riess, {\it et~al.\/}, {\it The Astronomical Journal\/} {\bf 116}, 1009
  (1998).

\bibitem{perlmutter_measurements_1999}
S.~Perlmutter, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 517}, 565
  (1999).

\bibitem{Riess_large_2019}
A.~G. Riess, S.~Casertano, W.~Yuan, L.~M. Macri, D.~Scolnic, {\it The
  Astrophysical Journal\/} {\bf 876}, 85 (2019).

\bibitem{amendola_cosmology_2013}
L.~Amendola, {\it et~al.\/}, {\it Living Reviews in Relativity\/} {\bf 16}, 6
  (2013).

\bibitem{spergel_wide_2015}
D.~Spergel, {\it et~al.\/}, {\it arXiv e-prints\/} {\bf arXiv:1503.03757}
  (2015).

\bibitem{Ivezic_lsst_2019}
Ivezic, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 873}, 111
  (2019).

\bibitem{aghanim_planck_2018}
{Planck Collaboration}, {\it et~al.\/}, {\it arXiv e-prints\/} {\bf
  arXiv:1807.06209} (2018).

\bibitem{verde_tensions_2019}
L.~Verde, T.~Treu, A.~G. Riess, {\it Nature Astronomy\/} {\bf 3}, 891 (2019).

\bibitem{einstein_lens_1936}
A.~Einstein, {\it Science\/} {\bf 84}, 506 (1936).

\bibitem{refsdal_possibility_1964}
S.~Refsdal, {\it Monthly Notices of the Royal Astronomical Society\/} {\bf
  128}, 307 (1964).

\bibitem{coe_cosmological_2009}
D.~Coe, L.~A. Moustakas, {\it The Astrophysical Journal\/} {\bf 706}, 45
  (2009).

\bibitem{linder_lensing_2011}
E.~V. Linder, {\it Physical Review D\/} {\bf 84}, 123529 (2011).

\bibitem{suyu_cosmology_2014}
S.~H. Suyu, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 788}, L35
  (2014).

\bibitem{bonvin_cosmograil_2019}
V.~Bonvin, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 629}, A97
  (2019).

\bibitem{wong_h0licow_2019}
K.~C. Wong, {\it et~al.\/}, {\it arXiv e-prints\/} {\bf arXiv:1907.04869}
  (2019).

\bibitem{suyu_cosmological_2018}
S.~H. Suyu, T.-C. Chang, F.~Courbin, T.~Okumura, {\it Space Science Reviews\/}
  {\bf 214}, 91 (2018).

\bibitem{kelly_multiple_2015}
P.~L. Kelly, {\it et~al.\/}, {\it Science\/} {\bf 347}, 1123 (2015).

\bibitem{goobar_iptf16geu:_2017}
A.~Goobar, {\it et~al.\/}, {\it Science\/} {\bf 356}, 291 (2017).

\bibitem{kelly_sn_2016}
P.~L. Kelly, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 831}, 205
  (2016).

\bibitem{dhawan_magnification_2019}
S.~Dhawan, {\it et~al.\/}, {\it Monthly Notices of the Royal Astronomical
  Society\/} 491, 2639 (2020).

\bibitem{newman_resolving_2018}
A.~B. Newman, S.~Belli, R.~S. Ellis, S.~G. Patel, {\it The Astrophysical
  Journal\/} {\bf 862}, 126 (2018).

\bibitem{ebeling_macs_2001}
H.~Ebeling, A.~C. Edge, J.~P. Henry, {\it The Astrophysical Journal\/} {\bf
  553}, 668 (2001).

\bibitem{materials_methods_2020}
Materials and methods are available as supplementary materials at the {Science}
  website.

\bibitem{jullo_bayesian_2007}
E.~Jullo, {\it et~al.\/}, {\it New Journal of Physics\/} {\bf 9}, 447 (2007).

\bibitem{kneib_lenstool_2011}
J.-P. Kneib, {\it et~al.\/} LENSTOOL: A Gravitational Lensing Software for Modeling Mass Distribution of Galaxies and Clusters, ascl:1102.004 (2011).

\bibitem{woosley_type_2007}
S.~E. Woosley, D.~Kasen, S.~Blinnikov, E.~Sorokina, {\it The Astrophysical
  Journal\/} {\bf 662}, 487 (2007).

\bibitem{kasen_type_2009}
D.~Kasen, S.~E. Woosley, {\it The Astrophysical Journal\/} {\bf 703}, 2205
  (2009).

\bibitem{oguri_gravitationally_2010}
M.~Oguri, P.~J. Marshall, {\it Monthly Notices of the Royal Astronomical
  Society\/} {\bf 405}, 2579 (2010).

\bibitem{goldstein_rates_2019}
D.~A. Goldstein, P.~E. Nugent, A.~Goobar, {\it The Astrophysical Journal
  Supplement Series\/} {\bf 243}, 6 (2019).

\bibitem{wojtak_magnified_2019}
R.~Wojtak, J.~Hjorth, C.~Gall, {\it Monthly Notices of the Royal Astronomical
  Society\/} {\bf 487}, 3342 (2019).

\bibitem{wang_determination_2006}
X.~Wang, L.~Wang, R.~Pain, X.~Zhou, Z.~Li, {\it The Astrophysical Journal\/}
  {\bf 645}, 488 (2006).

\bibitem{gonzaga_drizzlepac_2012}
S.~Gonzaga, {et al.}, {\it The {DrizzlePac} handbook\/}, HST Data Handbook (2012).

\bibitem{the_astropy_collaboration_astropy_2018}
{The Astropy Collaboration}, {\it et~al.\/}, {\it The Astronomical Journal\/}
  {\bf 156}, 123 (2018).

\bibitem{weilbacher_data_2020}
P.~M. Weilbacher, {\it et~al.\/}, {\it arXiv e-prints\/} {\bf arXiv:2006.08638}
  (2020).

\bibitem{bacon_muse_2017}
R.~Bacon, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 608}, A1
  (2017).

\bibitem{soto_zap_2016}
K.~T. Soto, S.~J. Lilly, R.~Bacon, J.~Richard, S.~Conseil, {\it Monthly Notices
  of the Royal Astronomical Society\/} {\bf 458}, 3210 (2016).

\bibitem{piqueras_mpdaf_2019}
L.~Piqueras, {\it et~al.\/}, {\it Astronomical data analysis software and
  systems {XXVI}\/}, M.~Molinaro, K.~Shortridge, F.~Pasian, eds. (2019), vol.
  521 of {\it Astronomical society of the pacific conference series\/}, p. 545.

\bibitem{mahler_strong_2018}
G.~Mahler, {\it et~al.\/}, {\it Monthly Notices of the Royal Astronomical
  Society\/} {\bf 473}, 663 (2018).

\bibitem{lagattuta_probing_2019}
D.~J. Lagattuta, {\it et~al.\/}, {\it Monthly Notices of the Royal Astronomical
  Society\/} {\bf 485}, 3738 (2019).

\bibitem{richard_mass_2014}
J.~Richard, {\it et~al.\/}, {\it Monthly Notices of the Royal Astronomical
  Society\/} {\bf 444}, 268 (2014).

\bibitem{richard_locuss_2010}
J.~Richard, {\it et~al.\/}, {\it Monthly Notices of the Royal Astronomical
  Society\/} {\bf 404}, 325 (2010).

\bibitem{mannucci_supernova_2005}
F.~Mannucci, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 433}, 807
  (2005).

\bibitem{newman_resolving_2018-1}
A.~B. Newman, S.~Belli, R.~S. Ellis, S.~G. Patel, {\it The Astrophysical
  Journal\/} {\bf 862}, 125 (2018).

\bibitem{foley_classifying_2013}
R.~J. Foley, K.~Mandel, {\it The Astrophysical Journal\/} {\bf 778}, 167
  (2013).

\bibitem{smartt_progenitors_2009}
S.~J. Smartt, {\it Annual Review of Astronomy and Astrophysics\/} {\bf 47}, 63
  (2009).

\bibitem{li_rates_2012}
X.~Li, J.~Hjorth, J.~Richard, {\it Journal of Cosmology and Astroparticle
  Physics\/} {\bf 2012}, 015 (2012).

\bibitem{barbary_sncosmo_2016}
K.~Barbary, {\it et~al.\/}, Sncosmo/{Sncosmo}: {V1}.4.0 (2016).

\bibitem{rodney_type_2014}
S.~A. Rodney, {\it et~al.\/}, {\it The Astronomical Journal\/} {\bf 148}, 13
  (2014).

\bibitem{guy_salt2:_2007}
J.~Guy, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 466}, 11 (2007).

\bibitem{pierel_extending_2018}
J.~D.~R. Pierel, {\it et~al.\/}, {\it Publications of the Astronomical Society
  of the Pacific\/} {\bf 130}, 114504 (2018).

\bibitem{kessler_snana:_2009}
R.~Kessler, {\it et~al.\/}, {\it Publications of the Astronomical Society of
  the Pacific\/} {\bf 121}, 1028 (2009).

\bibitem{frieman_sloan_2008}
J.~A. Frieman, {\it et~al.\/}, {\it The Astronomical Journal\/} {\bf 135}, 338
  (2008).

\bibitem{sako_sloan_2008}
M.~Sako, {\it et~al.\/}, {\it The Astronomical Journal\/} {\bf 135}, 348
  (2008).

\bibitem{dandrea_type_2010}
C.~B. D'Andrea, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 708},
  661 (2010).

\bibitem{astier_supernova_2006}
P.~Astier, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 447}, 31
  (2006).

\bibitem{hamuy_carnegie_2006}
M.~Hamuy, {\it et~al.\/}, {\it Publications of the Astronomical Society of the
  Pacific\/} {\bf 118}, 2 (2006).

\bibitem{stritzinger_he-rich_2009}
M.~Stritzinger, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 696},
  713 (2009).

\bibitem{morrell_carnegie_2012}
N.~I. Morrell, {\it Death of massive stars: {Supernovae} and gamma-ray
  bursts\/}, P.~Roming, N.~Kawai, E.~Pian, eds. (2012), vol. 279 of {\it {IAU}
  symposium\/}, pp. 361--362.

\bibitem{pierel_turning_2019}
J.~D.~R. Pierel, S.~Rodney, {\it The Astrophysical Journal\/} {\bf 876}, 107
  (2019).

\bibitem{huber_strongly_2019}
S.~Huber, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 631}, A161
  (2019).

\bibitem{goldstein_precise_2018}
D.~A. Goldstein, P.~E. Nugent, D.~N. Kasen, T.~E. Collett, {\it The
  Astrophysical Journal\/} {\bf 855}, 22 (2018).

\bibitem{foxley-marrable_impact_2018}
M.~Foxley-Marrable, T.~E. Collett, G.~Vernardos, D.~A. Goldstein, D.~Bacon,
  {\it Monthly Notices of the Royal Astronomical Society\/} {\bf 478}, 5081
  (2018).

\bibitem{bonvin_impact_2019}
V.~Bonvin, {\it et~al.\/}, {\it Astronomy \& Astrophysics\/} {\bf 621}, A55
  (2019).

\bibitem{kochanek_quantitative_2004}
C.~S. Kochanek, {\it The Astrophysical Journal\/} {\bf 605}, 58 (2004).

\bibitem{vernardos_joint_2018}
G.~Vernardos, {\it Monthly Notices of the Royal Astronomical Society\/} {\bf
  480}, 4675 (2018).

\bibitem{hsiao_k_2007}
E.~Y. Hsiao, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 663}, 1187
  (2007).

\bibitem{kessler_results_2010}
R.~Kessler, {\it et~al.\/}, {\it Publications of the Astronomical Society of
  the Pacific\/} {\bf 122}, 1415 (2010).

\bibitem{mosher_cosmological_2014}
J.~Mosher, {\it et~al.\/}, {\it The Astrophysical Journal\/} {\bf 793}, 16
  (2014).

\end{thebibliography}

Based on observations made with the NASA/ESA Hubble Space Telescope, obtained from the data archive at the Space Telescope Science Institute, and on observations collected at the European Organisation for Astronomical Research in the Southern Hemisphere under ESO programme 0103.A-0777(A).  
{\bf Funding:} Support for this work was provided by NASA through grant numbers HST-GO-15663 and HST-AR-15050 from the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Inc. under NASA contract NAS 5-26555.  The Cosmic Dawn Center of Excellence is funded by the Danish National Research Foundation under grant No. 140. This work was supported by NASA Headquarters under the NASA Future Investigators in Earth and Space Science and Technology (FINESST) award 80NSSC19K1414. 
{\bf Author Contributions:} 
Conceptualization, S.A.R., G.B., and S.T.; Methodology, S.A.R., J.D.R.P., J.R., and K.F.O.; Investigation, G.B., J.R., S.T., M.A., and K.E.W.; Writing - Original Draft, G.B.\ and S.T.; Writing - Review \& Editing, S.A.R., G.B., J.D.R.P., J.R., S.T., K.F.O., M.A., and K.E.W.; Visualization, S.A.R., G.B., J.D.R.P., J.R., K.F.O.; Supervision, G.B., S.T., and K.E.W.; Funding Acquisition, K.E.W.
{\bf Competing Interests:} All authors declare no competing interests.
{\bf Data and Materials Availability:} All HST images used in this work are available from the Mikulski Archive for Space Telescopes (mast.stsci.edu). All VLT MUSE spectroscopic data used in this work are available from  the ESO Archive Science Portal (archive.eso.org).  All high-level science products (photometry, lens model inputs, etc.) are available in the main text or the supplementary materials.  All software tools used for primary analysis tasks are publicly available, as indicated in the text.

@@

\clearpage
\section*{Supplementary materials}
Materials and Methods\\
Supplementary Text\\
Figures S1 to S8\\
Tables S1 to S7\\
References \textit{(32-68)}
\clearpage
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}

\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

\setcounter{page}{1}

\section*{Supplementary Materials for}
{\bf \large A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay}

>>>>

\documentclass[12pt]{article}
%\documentclass[12pt]{nature}

\usepackage[utf8]{inputenc}
%\usepackage[backend=biber,style=nature,citestyle=numeric]{biblatex}
%\usepackage{biblatex}
%\addbibresource{zotero_references_biblatex.bib}

@@

%\usepackage{amssymb}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage[draft,xcolor=usenames]{changes}
\definechangesauthor[name=1, color=blue]{r1}
\definechangesauthor[name=2, color=teal]{r2}
\definechangesauthor[name=3, color=orange]{r3}
\setauthormarkup{}

@@

%\usepackage{scicite}
%\usepackage{nature}

@@

%\paragraph*{One Sentence Summary:} A distant supernova has appeared as three images magnified by the gravity of a galaxy cluster, and will reappear in 20 years.

@@

  We report the discovery of a strongly-lensed supernova (SN) explosion that will enable a time delay measurement with an uncertainty of $<1\%$. This unprecedented scenario will enable a rare and highly precise measurement of the Hubble\added[id=r1]{-LeMa\^{i}tre} constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACS J0138.0-2155.  This is the most distant multiply-imaged supernova yet discovered.  We classify the supernova as a Type Ia with {$>90\%$} probability based on archival imaging observations from the \textit{Hubble Space Telescope}.  Three lensed images of the SN are detected with relative time delays of $<$200 days, and we predict that a fourth image in the component closest to the cluster core will appear in the year 2037$\pm$2.  Observing the light curve of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary 20 year baseline.

@@

%\section*{Main Text:}

Observations of Type Ia supernova (SN) explosions have played a key role in mapping the cosmic expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate 
\cite{riess_observational_1998,perlmutter_measurements_1999, Riess_large_2019}.  Determining the nature of dark energy and how it may evolve over time is a primary goal for the large-scale cosmology experiments of the 2020s
\cite{amendola_cosmology_2013,spergel_wide_2015,Ivezic_lsst_2019}.
A series of recent investigations into the expansion rate of the universe (the Hubble-LeMa\^itre constant; $H_0$) have found that measurements from the local universe are significantly different from the value inferred from measurement of the cosmic microwave background (CMB) radiation \cite{Riess_large_2019,aghanim_planck_2018}.  Resolving this tension in $H_0$ measurements may reveal new physics from the early universe, and to do so requires multiple 
independent cosmological probes \cite{verde_tensions_2019}.

One promising cosmological tool uses %time delays of 
gravitational lens systems in which a background
source appears as multiple images that arrive to the observer with relative delays \cite{einstein_lens_1936}.
%, which %for a given lens potential 
When such a strongly-lensed source is variable, one can measure the time delay between any pair of lensed images and derive a ratio of angular diameter distances to the foreground lens and background source. 
This distance ratio is sensitive to cosmological parameters, such as 
%the expansion rate of the universe, 
$H_0$  and the dark energy equation-of-state, $w$ \cite{refsdal_possibility_1964,coe_cosmological_2009,linder_lensing_2011}.
In recent years this method has been successfully applied to lensed
quasars \cite{suyu_cosmology_2014,bonvin_cosmograil_2019}, with six high-precision measurements to date \cite{wong_h0licow_2019}. As this sample of time delay lenses grows to several dozen, it is expected to deliver a measurement of $H_0$ with 1\% precision \cite{suyu_cosmological_2018}. \added[id=r1]{A future sample with $\sim100$ well-measured lensing time delays will be a competitive tool for dark energy studies \cite{linder_lensing_2011,jee_time-delay_2016,suyu_HOLISMOKES_2020}.}

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---in contrast to the stochastic variation of quasars.  However, despite the promise of lensed SN time delay cosmography as a cosmological distance probe, to date, there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_sn_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2019}. \added[id=r1]{An earlier discovery of an unusually luminous SN \cite{chornock_ps1-10afx_2013} was also shown to be a strongly-lensed SNIa, 
\cite{quimby_extraordinary_2013,quimby_detection_2014}
though the multiple images were not resolved.}  Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

@@

\added[id=r1]{By itself, \SNABC will not be particularly powerful as a solitary cosmological probe.  However,} future large-scale surveys such as the Vera C. Rubin Observatory 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their mission lifetimes, \added[id=r1]{though the vast majority of these will be lensed by galaxy-scale deflectors}  \cite{oguri_gravitationally_2010,goldstein_rates_2019,wojtak_magnified_2019}. \deleted[id=r1]{However, many of these lensed SN will require substantial follow-up resources to be made cosmologically useful.} Thus, almost all will have significantly shorter 
delays (of order 10--100 days) and therefore higher relative uncertainties than \SNABC.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like \SNABC can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by time delay measurement uncertainty.

In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN\added[id=r1]{---with $<1\%$ time delay precision and the same redshift ratio.} 
For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of \SNABC, the constraints 
from the \reqlike sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \reqlike sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \reqlike contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.

@@

\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout_2021.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the \SNABC images  (SN1--3) and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  
    \added[id=r1]{Labels marked 3.1--3.4 indicate the locations of a separate  multiply-imaged [O\textsc{ii}]-emitter at $z=0.77$ used to help constrain the lensing potential.} 
    Panels \emph{b}--\emph{i} show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels \emph{bcde} show the imaging from 2016 July where the SN was visible and \emph{fghi} show the later imaging from 2019 July where the the SN has faded away.  Panels \emph{aei} indicate the location of the fourth image predicted to appear in $\approx$2037. All panels use the late-epoch F125W imaging for the green channel; nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which helps to constrain the relative age of each SN image and the transient classification as a likely Type Ia supernova explosion. }
    \label{fig:layout}
\end{figure*}
\clearpage
\begin{figure}
  \centering
    \includegraphics[draft=False, width=\textwidth]{Paper/Figures/classification_contours_timeline.pdf}
    \caption{Classification information for SN Requiem based on its position in color-magnitude space. \emph{Left:} Contours show the population distributions for normal SNe of Type Ia (red), Type Ib/Ic (gold), and Type II (green), drawn to enclose 68\% (interior solid lines) and 95\% (exterior shaded regions) of each SN population.  Each SN sub-class was simulated at $z=1.95$, and samples from their expected light curves were drawn uniformly in time. \added[id=r1]{The directly observed photometry for each of the three SN images is shown as open markers. Vertical dotted lines show the effect of applying lens magnification corrections, with a filled grey marker indicating the corrected (demagnified) magnitude after applying our preferred lens model. Error bars indicate the systematic uncertainty (the range of alternative magnification corrections encompassed by lens model variations).}  \emph{Right:} The evolution of a typical Type Ia SN at $z=1.95$ in observed color-brightness space is shown by a colored line, with the line color indicating SN age in observer-frame days relative to peak brightness.  White diamonds correspond to the times labeled on the colorbar at right. Grey shading shows the typical range of luminosities observed for the Type Ia SN population in the nearby universe \cite{wang_determination_2006}. 
    %An interactive 3-D version of this figure is available online at \href{https://plot.ly/~jpierel/6}{https://plot.ly/~jpierel/6}
    \label{fig:class}
    }
\end{figure}

@@

\midrule
\textit{1}  & $87^{+16}_{-13}$ & -- & $5\pm1$ & --\\
\textit{2} & $-20^{+20}_{-11}$ & $106^{+25}_{-24}$ & $7\pm3$ & $82\pm62$ \\
\textit{3} & $100^{+27}_{-31}$ & $-10\pm24$ &$3.9\pm0.5$  & $-19\pm34$ \\
\textit{4} & -- & -- & $0.4\pm0.2$ & $7742\pm540$\\

%%%^^^^^^^UPDATE UNCERTAINTIES BASED ON JR^^^^^^^^%%%%%%

@@

@@

\clearpage
%\section*{Supplementary materials}
%Materials and Methods\\
%Supplementary Text\\
%Figures S1 to S8\\
%Tables S1 to S7\\
%References \textit{(32-68)}
%\clearpage
%\setcounter{table}{0}
%\renewcommand{\thetable}{S\arabic{table}}

%\setcounter{figure}{0}
%\renewcommand{\thefigure}{S\arabic{figure}}

%\setcounter{page}{1}

%\section*{Supplementary Materials for}
%{\bf \large A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay}

%\author{
% Steven A. Rodney$^{1\ast}$, 
% Gabriel Brammer$^{2\ast}$,  
% Justin D.\,R.~Pierel$^{1}$, 
% Johan Richard$^{3}$, 
% Sune Toft$^{2}$,  
% Kyle F. O'Connor$^{1}$, 
% Mohammad Akhshik$^{4}$, 
% Katherine E. Whitaker$^{2,4,5}$  \\
%\normalsize{$^{1}$Department of Physics \& Astronomy, University of South Carolina,} \normalsize{ Columbia, SC 29208, USA}\\
%\normalsize{$^{2}$Cosmic Dawn Center, Niels Bohr Institute, University of Copenhagen,} \normalsize{Lyngbyvej 2, Copenhagen \O\ 2100 Denmark}\\
%\normalsize{$^{3}$Univ Lyon, Univ Lyon1, Ens de Lyon, CNRS, Centre de Recherche Astrophysique de Lyon \normalsize{UMR5574, F-69230, Saint-Genis-Laval, France}\\
%\normalsize{$^{4}$Department of Physics, University of Connecticut,} \normalsize{Storrs, CT 06269, USA}\\
%\normalsize{$^{5}$Department of Astronomy, University of Massachusetts,} \normalsize{Amherst, MA 01003, USA}\\
%\normalsize{$^\ast$To whom correspondence should be addressed;}\\ \normalsize{E-mail:  srodney@sc.edu, gabriel.brammer@nbi.ku.dk}
%}

%\subsection*{This PDF file includes:}
